Modal Abstractions for Virtualizing Memory Addresses

### 2024-07-01 examples.tex
+

$\specline{ \textsf{entry-KERNBASE} \mapsto_{\textsf{vpte,qfrac}} \textsf{entry entry\_val} \ast \ulcorner  \textsf{qfrac} = 1 \leftrightarrow \; \lnot(\textsf{entry\_present entry\_val})\urcorner}$

@@

$\specline{ \textsf{entry-KERNBASE} \mapsto_{\textsf{vpte,qfrac}} \textsf{entry entry\_val} \ast \ulcorner  \textsf{qfrac} = 1 \leftrightarrow \; \lnot(\textsf{entry\_present entry\_val})  \urcorner }$

@@

$\specline{ \textsf{entry-KERNBASE} \mapsto_{\textsf{vpte,qfrac}} \textsf{entry entry\_val} \ast \ulcorner  \textsf{qfrac} = 1 \land \lnot(\textsf{entry\_present entry\_val})  \urcorner }$

@@

$\specline{ \textsf{entry-KERNBASE} \mapsto_{\textsf{vpte,qfrac}} \textsf{entry entry\_val} \ast \ulcorner  \textsf{qfrac} = 1 \land \lnot(\textsf{entry\_present entry\_val})  \urcorner }$
$\specline{ \texttt{rbp-16} \mapsto_{\textsf{v}} \textsf{nextpaddr} \ast \textsf{nextpaddr} \mapsto_{\textsf{p}} \textsf{w64\_zero}  \ast \texttt{rax} \mapsto_{\textsf{r}} \textsf{nextpaddr}  }$

@@

$\specline{ \textsf{entry-KERNBASE} \mapsto_{\textsf{vpte,qfrac}} \textsf{entry entry\_val} \ast \ulcorner  \textsf{qfrac} = 1 \land (\textsf{entry\_present entry\_val}) \urcorner }$
$\specline{ \texttt{rbp-16} \mapsto_{\textsf{v}} \textsf{nextpaddr}  \ast \texttt{rax} \mapsto_{\textsf{r}} \textsf{nextpaddr}  }$

@@

mov    -0x18[rbp],rax

@@

@@

As we see in Figure \ref{fig:calltopteinitialize}, we have the physical address (\textsf{entry}) of the page table entry kept in the register $\texttt{r8}$, but not the virtual memory mapping to load the entry value referenced by the physical address \textsf{entry} (Line 7 in Figure \ref{fig:calltopteinitialize}). We need to construct this virtual mapping for \textsf{entry}.
\subsubsection{Identity Mappings}
Most kernels maintain an invariant on their page tables that the virtual address of any page used for a page table lives at a virtual address
whose value is a constant offset from the physical address a practice sometimes referred to as \emph{identity mapping} (even though the physical-to-virtual translation
is typically not literally the identity function, but adding a constant offset).\footnote{Some kernels do this for all physical memory on the machine, simplifying interaction
with DMA devices.}
\begin{figure*}
\footnotesize
\[
\begin{array}{l}
  \mathcal{I}\textsf{ASpace}_{\textsf{id}}(\ptablestore,\Xi,m)\stackrel{\triangle}{=}\\ \textsf{ASpace\_Lookup}_{\textsf{id}}(\ptablestore,\Xi,m) \ast 
  \bigast{(\vaddr, \textsf{paddr})\in \ptablestore}{\exists\;(\textsf{l4e, l3e, l2e, l1e, paddr})\ldotp \textsf{L}_{4}\_\textsf{L}_{1}\_\textsf{PointsTo}(\vaddr\textsf{, l4e, l3e, l2e, l1e, paddr})} \\
  \bigast{(\paddr) \in \Xi}{\exists\; (\textsf{qfrac, q, val,}\vaddr) \ldotp \ulcorner \vaddr = \paddr + \textsf{OFF} \urcorner \ast  \underbrace{\sumwalkabs\vaddr\qfrac\paddr }_\text{Ghost translation} \ast 
% \qquad\qquad\qquad\qquad\qquad\qquad\qquad\qquad    
\underbrace{\paddr \mapsto_{\mathsf{p}}\{\textsf{qfrac}\}\; \vale}_\text{Physical location} \ast \ulcorner \textsf{qfrac} = 1 \leftrightarrow \; \lnot\textsf{entry\_present }(\vale) \urcorner }\\
  \textsf{ where } \\
   \textsf{ASpace\_Lookup}_{\textsf{id}}(\ptablestore,\Xi,m) \stackrel{\triangle}{=} \lambda\textsf{ cr3val} \ldotp \; \exists \gammaPred \; \ldotp \ulcorner m \; !!\; \textsf{cr3val} = \textsf{Some } \gammaPred \urcorner \ast
   \ownGhost\gammaPred{\authfull{\ptableabswalk\ptablestore}} \ast  \ownGhostpv\gammaPred{\authfull{\pvmapping\Xi}}

@@

Concretely speaking, before Line 7 in Figure \ref{fig:calltopteinitialize}, to read the value referenced by physical address \textsf{entry}, first we open our the addresss-space invariant ($\mathcal{I}\textsf{ASpace}_{\textsf{id}}(\ptablestore,\Xi,m)$), and then obtain the virtual mapping out of the identity mapping $\Xi$ with the side condition (\tag{*})
\begin{equation*}
 \ulcorner \textsf{qfrac} = 1 \leftrightarrow \; \lnot\textsf{entry\_present }(\vale) \urcorner \tag{*}
\end{equation*}
 assuring that looking up identity mappings are safe, because the full ownership asserts the non-presence of the entry which requires allocation a new physical page and setting the presence bit. By obtaining the identity mapping for \textsf{entry} in Figure \ref{fig:calltopteinitialize} -- ($\textsf{entry-KERNBASE} \mapsto_{\textsf{vpte,qfrac}} \textsf{entry entry\_val}$ with page addr \textsf{entry}),
 \[
 \begin{array}{l}
\specline{\textsf{P} \ast \mathcal{I}\texttt{ASpace}_{\textsf{id}}(\theta,\Xi\setminus\{entry\},m) \ast \textsf{rbp-8} \mapsto_{\textsf{v}} \textsf{entry} \ast r8  \mapsto_{\textsf{r}} \textsf{entry} \textsf{\_} \ast rdi  \mapsto_{\textsf{r}} \textsf{entry} \ast \textsf{entry-KERNBASE} \mapsto_{\textsf{vpte,qfrac}} \textsf{entry\_val} \ast \sumapaces\rtv\delta  \ast \sumpv\paddr\qfrac\true  }_{\rtv}
mov    [r8],rdi
\specline{\textsf{P} \ast \mathcal{I}\texttt{ASpace}_{\textsf{id}}(\theta,\Xi,m) \ast \textsf{rbp-8} \mapsto_{\textsf{v}} \textsf{entry} \ast r8  \mapsto_{\textsf{r}} \textsf{entry} \textsf{\_} \ast rdi  \mapsto_{\textsf{r}} \textsf{entry\_val} \ast  \textsf{entry-KERNBASE} \mapsto_{\textsf{vpte,qfrac}} \textsf{entry\_val} \ast \sumapaces\rtv\delta  \ast \sumpv\paddr\qfrac\true  }_{\rtv}
\end{array}
 \]
 we are able to load the \textsf{entry\_val} into \textsf{rdi}, and check the presence bit via $and \;0x1,rdi$.

 \subsubsection{Ensuring Presence of an Entry}
 Accessing the presence bit and checking the value allow (Lines 7-13 in Figure \ref{fig:calltopteinitialize}) us to exploit the fact in the side condition (*) in such a way that the condition ($\textsf{entry}\mapsto\textsf{present}$ determines whether we take the non-presence path (Lines 12-19 in Figure \ref{calltopteinitialize}) with full ownership ($\textsf{qfrac} = 1$) to set the entry with proper values for the bitfields, e.g., frame showing freshly allocated page and presence bit.

 In case of absence, \textsf{pte\_initialize} is called for a physical-zeroed page (utilizing the page-allocator's \textsf{kalloc} -- currently the only axiom as it is not part of \textsf{pte} (virtual-memory-management) module) to be backing the physical frame address of the entry.

 Once we make sure that the frame of the page-table entry ($\textsf{entry}\mapsto\textsf{pfn}$)is backed with an allocated physical page, we can obtain the virtual address of the next level-page table. To do so, again, we refer to the  (\textsf{next} of function \textsf{get\_next\_table} in Figure \ref{}) via looking up the \textit{identity mapping} from the physical address (the address value referenced by $\textsf{entry}\mapsto\textsf{pfn}$) to virtual address value to be returned by \textsf{get\_next\_table}.

 As part of the full page table walk verification, we have verified the critical step for a small x86-64 kernel, which is the physical-to-virtual
conversion (often appearing as a macro \texttt{P2V} inside the macro \textsf{ PTE\_PFN\_TO\_ADDR} in C source code). In our small kernel, as in larger kernels, \texttt{P2V} is actually just addition,
but the correctness of this simple instruction is quite subtle, as shown in Figure \ref{fig:p2v} -- showing the section of the \textsf{get\_next\_table} in Figure \ref{calltopteinitialize} after the entry presence ensured (Lines 1-19 in Figure \ref{fig:calltopteinitialize}).

\begin{figure}\footnotesize
\begin{lstlisting}[mathescape]
;;if (!entry->present) {
  ...

;;}
...
$\specline{\textsf{P} \ast \mathcal{I}\texttt{ASpace}_{\textsf{id}}(\theta,\Xi,m)  \ast \texttt{rbp-8} \mapsto_{\textsf{v}} \textsf{entry} \ast \texttt{r8}  \mapsto_{\textsf{r}} \textsf{\_} \ast \texttt{rdi}  \mapsto_{\textsf{r}} \textsf{\_} \ast \sumpv\paddr\qfrac\true  \sumapaces\rtv\delta  }_{\rtv}$ 
mov    rax,-0x18[rbp]

;;uintptr_t next_virt_addr = (uintptr_t) P2V(next_phys_addr);
mov    -0x18[rbp],rax
movabs 0x0,rcx            187: R_X86_64_64	KERN_BASE
add    rcx,rax
mov    rax,-0x20[rbp]
\end{lstlisting}
\vspace{-1em}
\caption{Converting a physical address, whose exsitence is made certain, of a PTE to a virtual address (w/o instruction pointer or flag updates). $R$ abbreviates the resources
associated with \paddr\ by $\Xi$ in Figure \ref{fig:peraspaceinvariant_with_p2v_extension}.}
\label{fig:p2v}
\end{figure}

>>>>

$\specline{ \textsf{entry+KERNBASE} \mapsto_{\textsf{vpte,qfrac}} \textsf{entry entry\_val} \ast \ulcorner  \textsf{qfrac} = 1 \leftrightarrow \; \lnot(\textsf{entry\_present entry\_val})\urcorner}_{\rtv}$

@@

$\specline{ \textsf{entry+KERNBASE} \mapsto_{\textsf{vpte,qfrac}} \textsf{entry entry\_val} \ast \ulcorner  \textsf{qfrac} = 1 \leftrightarrow \; \lnot(\textsf{entry\_present entry\_val})  \urcorner }_{\rtv}$

@@

$\specline{ \textsf{entry+KERNBASE} \mapsto_{\textsf{vpte,qfrac}} \textsf{entry entry\_val} \ast \ulcorner  \textsf{qfrac} = 1 \land \lnot(\textsf{entry\_present entry\_val})  \urcorner }_{\rtv}$

@@

$\specline{ \textsf{entry+KERNBASE} \mapsto_{\textsf{vpte,qfrac}} \textsf{entry entry\_val} \ast \ulcorner  \textsf{qfrac} = 1 \land v\lnot(\textsf{entry\_present entry\_val})  \urcorner }_{\rtv}$
$\specline{ \texttt{rbp-16} \mapsto_{\textsf{v}} \textsf{nextpaddr} \ast \textsf{nextpaddr} \mapsto_{\textsf{p}} \textsf{w64\_zero}  \ast \texttt{rax} \mapsto_{\textsf{r}} \textsf{nextpaddr}  }_{\rtv}$

@@

$\specline{ \textsf{entry+KERNBASE} \mapsto_{\textsf{vpte,qfrac}} \textsf{entry entry\_val} \ast \ulcorner  \textsf{qfrac} = 1 \land (\textsf{entry\_present entry\_val}) \urcorner }_{\rtv}$
$\specline{ \texttt{rbp-16} \mapsto_{\textsf{v}} \textsf{nextpaddr}  \ast \texttt{rax} \mapsto_{\textsf{r}} \textsf{nextpaddr}  }_{\rtv}$

@@

@@

;;next = (pte_t *) next_virt_addr;
;;clean up the stack and return next

@@

As we see in Figure \ref{fig:calltopteinitialize}, we have the physical address (\textsf{entry}) of the page table entry kept in the register $\texttt{r8}$, but not the virtual memory mapping to load the entry value referenced by the physical address \textsf{entry} (Line 7 in Figure \ref{fig:calltopteinitialize}). We need to construct this virtual mapping for \textsf{entry}.
\subsubsection{Identity Mappings}
\label{subsec:identitymappings}
Most kernels maintain an invariant on their page tables that the virtual address of any page used for a page table lives at a virtual address
whose value is a constant offset from the physical address a practice sometimes referred to as \emph{identity mapping} (even though the physical-to-virtual translation
is typically not literally the identity function, but adding a constant offset).\footnote{Some kernels do this for all physical memory on the machine, simplifying interaction
with DMA devices.}
\begin{figure*}
\footnotesize
\[
\begin{array}{l}
  \mathcal{I}\textsf{ASpace}_{\textsf{id}}(\ptablestore,\Xi,m)\stackrel{\triangle}{=}\\ \textsf{ASpace\_Lookup}_{\textsf{id}}(\ptablestore,\Xi,m) \ast 
  \bigast{(\vaddr, \textsf{paddr})\in \ptablestore}{\exists\;(\textsf{l4e, l3e, l2e, l1e, paddr})\ldotp \textsf{L}_{4}\_\textsf{L}_{1}\_\textsf{PointsTo}(\vaddr\textsf{, l4e, l3e, l2e, l1e, paddr})} \\
  \bigast{(\paddr) \in \Xi}{\exists\; (\textsf{qfrac, q, val,}\vaddr) \ldotp \ulcorner \vaddr = \paddr + \textsf{OFF} \urcorner \ast  \underbrace{\sumwalkabs\vaddr\qfrac\paddr }_\text{Ghost translation} \ast 
% \qquad\qquad\qquad\qquad\qquad\qquad\qquad\qquad    
\underbrace{\paddr \mapsto_{\mathsf{p}}\{\textsf{qfrac}\}\; \vale}_\text{Physical location} \ast \ulcorner \textsf{qfrac} = 1 \leftrightarrow \; \lnot\textsf{entry\_present }(\vale) \urcorner }\\
  \textsf{ where } \\
   \textsf{ASpace\_Lookup}_{\textsf{id}}(\ptablestore,\Xi,m) \stackrel{\triangle}{=} \lambda\textsf{ cr3val} \ldotp \; \exists \gammaPred \; \ldotp \ulcorner m \; !!\; \textsf{cr3val} = \textsf{Some } \gammaPred \urcorner \ast
   \ownGhost\gammaPred{\authfull{\ptableabswalk\ptablestore}} \ast  \ownGhostpv\gammaPred{\authfull{\pvmapping\Xi}}

@@

Concretely speaking, before Line 7 in Figure \ref{fig:calltopteinitialize}, to read the value referenced by physical address \textsf{entry}, first we open our the addresss-space invariant ($\mathcal{I}\textsf{ASpace}_{\textsf{id}}(\ptablestore,\Xi,m)$), and then obtain the virtual mapping out of the identity mapping $\Xi$ with the side condition (*)
\begin{equation*}
 \ulcorner \textsf{qfrac} = 1 \leftrightarrow \; \lnot\textsf{entry\_present }(\vale) \urcorner \tag{*}
\end{equation*}
 assuring that looking up identity mappings are safe, because the full ownership asserts the non-presence of the entry which requires allocation a new physical page and setting the presence bit. By obtaining the identity mapping for \textsf{entry} in Figure \ref{fig:calltopteinitialize} -- ($\textsf{entry+KERNBASE} \mapsto_{\textsf{vpte,qfrac}} \textsf{entry entry\_val}$ with page addr \textsf{entry}),
 \[
 \begin{array}{l}
\specline{\textsf{P} \ast \mathcal{I}\texttt{ASpace}_{\textsf{id}}(\theta,\Xi\setminus\{entry\},m) \ast \textsf{rbp-8} \mapsto_{\textsf{v}} \textsf{entry} \ast r8  \mapsto_{\textsf{r}} \textsf{entry} \textsf{\_} \ast rdi  \mapsto_{\textsf{r}} \textsf{entry} \ast \textsf{entry+KERNBASE} \mapsto_{\textsf{vpte,qfrac}} \textsf{entry\_val} \ast \sumapaces\rtv\delta  \ast \sumpv\paddr\qfrac\true  }_{\rtv}
mov    [r8],rdi
\specline{\textsf{P} \ast \mathcal{I}\texttt{ASpace}_{\textsf{id}}(\theta,\Xi,m) \ast \textsf{rbp-8} \mapsto_{\textsf{v}} \textsf{entry} \ast r8  \mapsto_{\textsf{r}} \textsf{entry} \textsf{\_} \ast rdi  \mapsto_{\textsf{r}} \textsf{entry\_val} \ast  \textsf{entry+KERNBASE} \mapsto_{\textsf{vpte,qfrac}} \textsf{entry\_val} \ast \sumapaces\rtv\delta  \ast \sumpv\paddr\qfrac\true  }_{\rtv}
\end{array}
 \]
 we are able to load the \textsf{entry\_val} into \textsf{rdi}, and check the presence bit via $and \;0x1,rdi$.

 \subsubsection{Ensuring Presence of an Entry}
 Accessing the presence bit and checking the value allow (Lines 7-13 in Figure \ref{fig:calltopteinitialize}) us to exploit the fact in the side condition (*) in such a way that the condition ($\textsf{entry}\mapsto\textsf{present}$ determines whether we take the non-presence path (Lines 12-19 in Figure \ref{fig:calltopteinitialize}) with full ownership ($\textsf{qfrac} = 1$) to set the entry with proper values for the bitfields, e.g., frame showing freshly allocated page and presence bit.
  In case of absence, \textsf{pte\_initialize} is called for a physical-zeroed page (utilizing the page-allocator's \textsf{kalloc} -- currently the only axiom as it is not part of \textsf{pte} (virtual-memory-management) module) to be backing the physical frame address of the entry -- Specification Line 35 in Figure \ref{fig:calltopteinitialize}.

  \subsubsection{Kernels Utilizing \textsf{P2V}}
  \label{sec:p2v}
 Once we make sure that the frame of the page-table entry ($\textsf{entry}\mapsto\textsf{pfn}$)is backed with an allocated physical page, we must obtain the virtual address of the next level-page table (Line 46 in \ref{calltopteinitialize}). To do so, as for obtaining any virtual page-table address, we, again, need to open the invariant look up the \textit{identity mapping} for the physical address ($\textsf{entry}\mapsto \textsf{pfn}$).

This actually constitutes a very critical piece of the full page table walk verification, and we have verified the critical step for a small x86-64 kernel, which is the physical-to-virtual conversion (often appearing as a macro \texttt{P2V} inside the macro \textsf{ PTE\_PFN\_TO\_ADDR} in C source code). In our small kernel, as in larger kernels, \texttt{P2V} is actually just addition, but the correctness of this simple instruction is quite subtle and relies on identity mappings explained in the previous section (Section \ref{subsec:identitymappings}) as a part of address-space invariant.  As as separate Lemma shown in Figure \ref{fig:p2v}, 
\begin{figure}\footnotesize
\begin{lstlisting}[mathescape]
...
;;uintptr_t next_virt_addr = (uintptr_t) P2V(next_phys_addr);
movabs 0x0,rcx	                187: R_X86_64_64	KERN_BASE
$\specline{\textsf{P} \ast \mathcal{I}\texttt{ASpace}_{\textsf{id}}(\theta,\Xi,m)  \ast \textsf{rbp-8} \mapsto_{\textsf{v}} \textsf{entry} \ast \texttt{r8}  \mapsto_{\textsf{r}} \textsf{entry}  \ast \texttt{rdi}  \mapsto_{\textsf{r}} \textsf{entry\_val}  }_{\rtv}$
$\specline{\sumpv\paddr\qfrac\true \ast \sumapaces\rtv\delta  }_{\rtv}$
$\specline{ \textsf{entry+KERNBASE} \mapsto_{\textsf{vpte,qfrac}} \textsf{entry entry\_val} \ast \ulcorner  \textsf{qfrac} = 1 \land (\textsf{entry\_present entry\_val}) \urcorner }$
$\specline{ \texttt{rbp-16} \mapsto_{\textsf{v}} \textsf{nextpaddr}  \ast \texttt{rax} \mapsto_{\textsf{r}} \textsf{nextpaddr} \ast  \texttt{rcx} \mapsto_{\textsf{r}} \textsf{KERNBASE}  }_{\rtv}$
add    rcx,rax
$\specline{\textsf{P} \ast \mathcal{I}\texttt{ASpace}_{\textsf{id}}(\theta,\Xi,m)  \ast \textsf{rbp-8} \mapsto_{\textsf{v}} \textsf{entry} \ast \texttt{r8}  \mapsto_{\textsf{r}} \textsf{entry}  \ast \texttt{rdi}  \mapsto_{\textsf{r}} \textsf{entry\_val}  }_{\rtv}$
$\specline{\sumpv\paddr\qfrac\true \ast \sumapaces\rtv\delta  }_{\rtv}$
$\specline{ \textsf{entry+KERNBASE} \mapsto_{\textsf{vpte,qfrac}} \textsf{entry entry\_val} \ast \ulcorner  \textsf{qfrac} = 1 \land (\textsf{entry\_present entry\_val}) \urcorner }$
$\specline{ \texttt{rbp-16} \mapsto_{\textsf{v}} \textsf{nextpaddr}  \ast \texttt{rax} \mapsto_{\textsf{r}} \textsf{nextpaddr} \ast \texttt{rcx} \mapsto_{\textsf{r}} \textsf{KERNBASE} }_{\rtv}$
$\specline{ \texttt{nextpaddr+KERNBASE} \mapsto_{\textsf{vpte,qfracn}} \textsf{nextpaddr w64\_zero} \ }_{\rtv}$
...
;;next = (pte_t *) next_virt_addr;
;;clean up the stack and return next
\end{lstlisting}
\vspace{-1em}
\caption{Converting a physical address of a PTE to a virtual address (w/o instruction pointer or flag updates). $R$ abbreviates the resources
associated with \paddr\ by $\Xi$ in Figure \ref{fig:peraspaceinvariant_with_p2v_extension}.}
\label{fig:p2v}
\end{figure}

%\begin{figure}\footnotesize
%\begin{lstlisting}[mathescape]
%;;if (!entry->present) {
%  ...
%
%;;}
%...
%$\specline{\textsf{P} \ast \mathcal{I}\texttt{ASpace}_{\textsf{id}}(\theta,\Xi,m)  \ast \texttt{rbp-8} \mapsto_{\textsf{v}} \textsf{entry} \ast \texttt{r8}  \mapsto_{\textsf{r}} \textsf{\_} \ast \texttt{rdi}  \mapsto_{\textsf{r}} \textsf{\_} \ast \sumpv\paddr\qfrac\true  \sumapaces\rtv\delta  }_{\rtv}$ 
%mov    rax,-0x18[rbp]
 % 
%;;uintptr_t next_virt_addr = (uintptr_t) P2V(next_phys_addr);
%mov    -0x18[rbp],rax
%movabs 0x0,rcx            187: R_X86_64_64	KERN_BASE
%add    rcx,rax
%mov    rax,-0x20[rbp]
%\end{lstlisting}
%\vspace{-1em}
%\caption{Converting a physical address, whose exsitence is made certain, of a PTE to a virtual address (w/o instruction pointer or flag updates). $R$ abbreviates the resources
%associated with \paddr\ by $\Xi$ in Figure \ref{fig:peraspaceinvariant_with_p2v_extension}.}
%\label{fig:p2v}
%\end{figure}
